---
title: "Assignment 5: Data Visualization"
author: "Student Name"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1
library(tidyverse)
library(here)
library(cowplot)
library(ggthemes)

here()


#2 
#reading in CSVs
lake_chem_proc <- 
  read.csv(here(
"Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
stringsAsFactors = TRUE)

litter_trap_proc <-
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
           stringsAsFactors = TRUE)


#checking the format for dates
class(lake_chem_proc$sampledate)
class(litter_trap_proc$collectDate)

#converting to date format
lake_chem_proc$sampledate <- ymd(lake_chem_proc$sampledate)
litter_trap_proc$collectDate <- ymd(litter_trap_proc$collectDate)


```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

A5_theme <- theme_base()  + 
  theme(
    rect = element_rect(
      fill = "ivory"
    ),
    plot.title =element_text(
      face = "bold",
      color = "royalblue1", 
      size = 12
    ),
    axis.text = element_text(
      family = "italic",
      size = 10
    ),
    axis.title = element_text(
      size = 10
    )
    )

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

plot4 <-  lake_chem_proc %>% 
  #filter(depth == 0) %>% 
  ggplot(aes(x = tp_ug, y = po4, color = lakename))+
  geom_point()+
  xlab("Total Phosphorus")+
  ylab("Phosphate")+
  ylim(0,50)+
  geom_smooth(method = lm)+
  labs(title = "Total Phosphorus vs Phosphate")+
  A5_theme
print(plot4)

```

5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
plot5_temp <- lake_chem_proc %>%
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = temperature_C, 
             color = lakename)) +
  geom_boxplot() +
  labs(x = "Month", 
       y = "Temperature (C)",
       title = "Temperature, TN, and TP by Month") +
  A5_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
print(plot5_temp)

plot5_tn <- lake_chem_proc %>%
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = tn_ug, 
             color = lakename)) +
  geom_boxplot() +
  labs(x = "Month", 
       y = "Total Nitrogen (ug)",
       title = "Temperature, TN, and TP by Month") +
  A5_theme +
   theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(), 
    plot.title = element_blank()
  )
print(plot5_tn)

plot5_tp <- lake_chem_proc %>%
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = tp_ug, 
             color = lakename)) +
  geom_boxplot() +
  labs(x = "Month", 
       y = "Total Phosphorus (ug)",
       title = "Temperature, TN, and TP by Month") +
  A5_theme +
  theme(
    plot.title = element_blank()
  )
print(plot5_tp)

season_vars <- plot_grid(plot5_temp, plot5_tn, plot5_tp, nrow = 3, align = 'h', rel_heights = c(1.25, 1))
print(season_vars)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

litter_all <- litter_trap_proc %>% 
  filter(functionalGroup == 'Needles') %>% 
   ggplot(aes(x = collectDate, 
             y = dryMass,
             color = nlcdClass))+
  geom_point()+
  A5_theme

print(litter_all)

#7

litter_comb <- litter_trap_proc %>% 
  filter(functionalGroup == 'Needles') %>% 
  ggplot(aes(x = collectDate, 
             y = dryMass)) +
  facet_wrap(facets = vars(nlcdClass), ncol = 3) +
  geom_point() +
  A5_theme

print(litter_comb)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: